Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent

### 2023-06-26 manuscript/main.tex

\section{Algorithm}

\paragraph{The HOGHWILD! path-guided stochastic gradient descent algorithm}
Our algorithm is inspired by Zheng et al. (\url{https://doi.org/hkv4}) and it is theoretically applicable in any number of dimensions.
In this work, we describe the results in 1D and 2D.
In our implementation, the algorithm moves a single pair of nodes at a time, minimizing the disparity between the layout distance of a node pair and the actual nucleotide distance of a path traversing these nodes.
Our path index enables efficient navigation of path positions by updating nodes on the fly, instead of keeping all pairwise distances of all nodes in memory.
This allows the PG-SGD to be applied manuscriptto gigabase-scale pangenome graphs. \\
Our multi-threaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
\begin{itemize}
	\item The first node Xi of a pair is a uniform path step pick from all nodes.
	\item The second node Xj of a pair is sampled from the same path following a Zipfian distribution. 
	\item The path nucleotide distance of the nodes in the pair guides the actual layout distance dij update of these nodes. The magnitude r of the update depends on the current learning rate of the SGD.
\end{itemize}
%The exploration of the path-guided SGD parameter space is in progress, in order to get the best layout as quickly as possible.
We describe in algorithm~\ref{alg:pg_sgd} how the PG-SGD works on a single thread. We can have several worker threads but only one checker thread!

@@

% something for the supplementary
\paragraph{Performance evaluation}
Fig. 2: Performance evaluation 1D + 2D: Time + RAM by number of haplotypes (Fig.~\ref{fig:haps_time_ram}). Time + RAM by number of threads (Fig.~\ref{fig:threads_time_ram}). Full HPRC graph.
\\
1D (max. threads) vs. ALIBI; we should compare time + RAM in Fig.~\ref{fig:alibi}.
But also by our sorting goodness metrics: odgi sort + the one suggested in the ALIBI paper in Table 1.
\\
2D (max. threads) + odgi draw vs. Bandage in Fig.~\ref{fig:bandage}.
\input{fig_performance}
\\
\\
Table 1: Metrics of the ALIBI and PG-SGD graphs from Fig. 2.
\input{tab_performance.tex}
Already, the PG-SGD outperforms existing graph linearization methods like the flow procedure (\url{https://doi.org/gdw58w}) or ALIBI (\url{https://doi.org/hkv3}).
\paragraph{Latent graph structure reveals underlying biology}
Fig. 3: Cool quantitative 1D sortings and 2D layouts: biological implications.
Randomly sorted. PG-SGD sorted. Ygs sorted. Reference sorted.
We want a pipeline of sortings. 2D layout of the whole HPRC.
Chr6 HPRC HLA graph? Chr8 beta-defensin gene cluster HPRC? Whole HPRC?
\\
We could also try to build a gastric cancer pangenome graph with data from \url{https://www.nature.com/articles/s41467-022-33073-7} from up to 185 samples.
However, we would have to request access to the data.
\\
\\
\input{fig_latent_graph_structure}
Table 2: Metrics of the sorted graphs in Fig. 3.
\input{tab_metrics_latent_graph_structure}
\paragraph{Bonus Section}
Fig. 4: Detect tension. Relax a graph. Detect tension afterwards.
I need to test this on a new data set I got from Erik.
I need to establish a fixed lower boundary for the tension from which on we don't relax anymore. \\
With a high quality layout, we can measure the discrepancy of the path layout position versus the expected path nucleotide position, the “tension” of a graph.
The greater the “tension”, the greater is the possibility of a biologically meaningless alignment.
This allows us to detect telomere collapsing alignment errors and hopefully (pangenome) assembly errors, subsequently correcting them.
\input{fig_tension}
\fi

@@

We propose / implemented ....
\paragraph{}
Difference to other existing methods, are there possible improvements of our method possible?
\paragraph{}
Discuss performance eval
@Jiajie + Niklas: What about going GPU?
\paragraph{}
The algorithm allows to inspect a pangenome graph on base-level, as a whole.
\paragraph{from the ODGI paper:}
Its static, large-scale 1D and 2D visualizations of the pangenome graphs allow an unprecedented high-level perspective on variation in pangenomes, and have also been critical in the development of pangenome graph building methods.
However, an interactive solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.
Recent interactive pangenome graph browsers are reference-centric (Beyer et al., 2019; Yokoyama et al., 2019), have a limited predefined coordinate system (Durant et al., 2021), or focus primarily on 2D representations (Gonnella et al., 2019; Wick et al., 2015).
Our graph sorting and layout algorithms can provide the foundation for future tools of this type.
We plan to focus on using these learned models to detect structural variation and assembly errors.
\paragraph{}
The graph simplification pipeline smoothxg runs POA for each block of paths that are co-linear within each seqwish induced variation graph.
A prerequisite is that the graph nodes are sorted according to their occurrence in the graph's embedded paths.
Our 1D path-guided SGD algorithm is designed to provide this kind of sort.
Already, the 1D PG-SGD is a key step in the PanGenome Graph Building (PGGB) pipeline that we successfully applied to build the first draft human pangenome reference (Liao et al., bioRxiv 2022).
\paragraph{}
I'd like to argue that this projection presents biologically relevant structures that we can work with downstream
both for human interface
and for other algorithms that look at and classify variation
\paragraph{}
What about the future?
\paragraph{}
How can other scientists benefit from this work?
Graph visualization (together with annotation) is key for understanding variations.
This projection presents biologically relevant structures that we can work with downstream for researches and algorithms that look at and classify variation,

>>>>

    \paragraph{The HOGHWILD! path-guided stochastic gradient descent algorithm}
    \FIXME{SPECIFY THAT THE IMPLEMENTATION IS IN ODGI AND CITE IT}
    Our algorithm is inspired by Zheng et al. ((\FIXME{CITATION} \url{https://doi.org/hkv4}) and it is applicable in any number of dimensions.
    In this work, we describe the results in 1 dimension (1D) and 2 dimensions (2D).
    In our implementation, the algorithm moves a single pair of nodes at a time, minimizing the disparity between the layout distance of a node pair and the actual nucleotide distance of a path traversing these nodes.
    Our path index enables efficient navigation of path positions by updating nodes on the fly, instead of keeping all pairwise distances of all nodes in memory.
    This allows the PG-SGD to be applied to gigabase-scale pangenome graphs. \\
    Our multithreaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
    \begin{itemize}
        \item The first node Xi of a pair is a uniform path step pick from all nodes.
        \item The second node Xj of a pair is sampled from the same path following a Zipfian distribution.
        \item The path nucleotide distance of the nodes in the pair guides the actual layout distance dij update of these nodes.
        \item The magnitude r of the update depends on the current learning rate of the SGD.
    \end{itemize}
%The exploration of the path-guided SGD parameter space is in progress, in order to get the best layout as quickly as possible.
    We describe in algorithm~\ref{alg:pg_sgd} how the PG-SGD works on a single thread.
    We can have several worker threads but only one checker thread!

@@

% something for the supplementary

    \paragraph{Performance evaluation}
    Fig. 2: Performance evaluation 1D + 2D: Time + RAM by number of haplotypes (Fig.~\ref{fig:haps_time_ram}). Time + RAM by number of threads (Fig.~\ref{fig:threads_time_ram}). Full HPRC graph.
    \\
    1D (max. threads) vs. ALIBI; we should compare time + RAM in Fig.~\ref{fig:alibi}.
    But also by our sorting goodness metrics: odgi sort + the one suggested in the ALIBI paper in Table 1.
    \\
    2D (max. threads) + odgi draw vs. Bandage in Fig.~\ref{fig:bandage}.
    \input{fig_performance}
    \\
    \\
    Table 1: Metrics of the ALIBI and PG-SGD graphs from Fig. 2.
    \input{tab_performance.tex}
    Already, the PG-SGD outperforms existing graph linearization methods like the flow procedure (\url{https://doi.org/gdw58w}) or ALIBI (\url{https://doi.org/hkv3}).

    \paragraph{Latent graph structure reveals underlying biology}
    Fig. 3: Cool quantitative 1D sortings and 2D layouts: biological implications.
    Randomly sorted. PG-SGD sorted. Ygs sorted. Reference sorted.
    We want a pipeline of sortings. 2D layout of the whole HPRC.
    Chr6 HPRC HLA graph? Chr8 beta-defensin gene cluster HPRC? Whole HPRC?
    \\
    We could also try to build a gastric cancer pangenome graph with data from \url{https://www.nature.com/articles/s41467-022-33073-7} from up to 185 samples.
    However, we would have to request access to the data.
    \\
    \\
    \input{fig_latent_graph_structure}
    Table 2: Metrics of the sorted graphs in Fig. 3.
    \input{tab_metrics_latent_graph_structure}

    \paragraph{Bonus Section}
    Fig. 4: Detect tension. Relax a graph. Detect tension afterwards.
    I need to test this on a new data set I got from Erik.
    I need to establish a fixed lower boundary for the tension from which on we don't relax anymore. \\
    With a high quality layout, we can measure the discrepancy of the path layout position versus the expected path nucleotide position, the “tension” of a graph.
    The greater the “tension”, the greater is the possibility of a biologically meaningless alignment.
    This allows us to detect telomere collapsing alignment errors and hopefully (pangenome) assembly errors, subsequently correcting them.
    \input{fig_tension}
    \fi

@@

    We presented the Path-Guided Stochastic Gradient Descent (PG-SGD), a layout algorithm for pangenome graphs that
    leverages the inherent biological information available within the graph's represented genomic sequences.
    Our implementation efficiently computes the layout of pangenome graphs representing thousands of individuals at whole-genome scale.
    \FIXME{Difference to other existing methods, are there possible improvements of our method possible?}
    \FIXME{Discuss performance eval. @Jiajie + Niklas: What about going GPU?}
    We implemented the PG-SGD algorithm to generate layouts in 1 and 2 dimensions, but the algorithm can be extended in any number of dimensions.
    Graph visualization is key for understanding genome variations and the large-scale layouts produced by the PG-SGD algorithm offer an unprecedented 
    high-level perspective on variation in pangenomes. These layouts are critical for pangenome graph building (\FIXME{CITE PGG and HPRC MAIN PAPER}) and 
    making new biological discoveries (\FIXME{CITE ACROCENTRIC PAPER}).
    %The graph simplification pipeline smoothxg runs POA for each block of paths that are co-linear within each seqwish induced variation graph.
    %A prerequisite is that the graph nodes are sorted according to their occurrence in the graph's embedded paths.
    %Our 1D path-guided SGD algorithm is designed to provide this kind of sort.
    %Already, the 1D PG-SGD is a key step in the PanGenome Graph Building (PGGB) pipeline that we successfully applied to build the first draft human pangenome reference (Liao et al., bioRxiv 2022).
    However, there remains a gap in interactive and scalable solutions that merge layouts of large pangenome graph with annotation.
    Our PG-SGD algorithm will be the foundation of new pangenome graph browsers for exploring meaningful graph layouts.
    Futhermore, projecting a graph in a low number of dimensions, but still preserving its biologically relevant information, enable the application of 
    algorithms that use the graph layout to look at variant classification.
    Our future research involve exploiting these learned projections to detect structural variations and assembly errors.
    \FIXME{PROTEIN?}
